---
title: "Homework on estimating dynamic discrete choice models"
date: "`r format(Sys.time(), '%d %B, %Y')`"
output:
  html_document:
    toc: true
    toc_depth: 3
    toc_float:
      collapsed: false
      smooth_scroll: false
---

```{r,echo=FALSE,message=FALSE,warning=FALSE}
require(data.table)
```

In this homework we want to estimate a simple dynamic discrete choice model that we saw in class. See also the paper [Keane and Wolpin](). They propose a simple model of female labor supply as follows. Utility for married woman $i$ in period $t$ from working (option 1) vs not working (option 0) with $n_i$ small children is

$$
\begin{align}
U_{it}^1 &= y_{it} + w_{it} - \pi n_{it} \\
U_{it}^0 &= y_{it} + x_{it}\beta + \epsilon_{it} 
\end{align}
$$
where $y_{it}$ is the husband's income. Let's write the difference in those utilities as $U_{it}^1 - U_{it}^0$,

$$
v_{it}(x_{it},w_{it},n_{it},\epsilon_{it}) = w_{it} - \pi n_{it} -x_{it}\beta - \epsilon_{it} 
$$
and define the work indicator $d_{it} = \mathbf{1}[U_{it}^1 > U_{it}^0]$. Next, we define the state space as observed by the econometrician as $\Omega_{it} = (x_{it},w_{it},n_{it})$, i.e. we don't observe $\epsilon$, but the decision maker does.

Woman $i$ will work in $t$ if $U_{it}^1 > U_{it}^0$, i.e. if $v_{it}(x_{it},w_{it},n_{it},\epsilon_{it}) > 0$, and at $v_{it}(x_{it},w_{it},n_{it},\epsilon_{it}) = 0$ she is indifferent. Call the $\epsilon$ that solves this *the critical epsilon* $\epsilon^*(\Omega_{it})$. We have that, given state $\Omega_{it}$,

$$
i\text{ chooses to }\begin{cases}\text{work in t}&\text{if }\epsilon_{it} < \epsilon^*(\Omega_{it}) \Rightarrow U_{it}^1 > U_{it}^0 \\
\text{not work in t}& \text{if }\epsilon_{it} > \epsilon^*(\Omega_{it}) \Rightarrow U_{it}^1 < U_{it}^0\end{cases}
$$

Assuming that $\epsilon$ is independent of $\Omega$, the probability of working is computed by looking at each $i$'s $\epsilon_{it}$ and couting whether it's below $\epsilon^*(\Omega_{it})$:

$$
\Pr[d_{it}=1|\Omega{it}] = \int_{-\infty}^{\epsilon_{it}^*} dF_{\epsilon}(\epsilon|\Omega{it}) = \int_{-\infty}^{\epsilon_{it}^*} dF_{\epsilon}(\epsilon)
$$

Clearly $\Pr[d_{it}=0|\Omega{it}] = 1- \Pr[d_{it}=1|\Omega{it}]$, and so the likelihood for a random sample of $N$ females observed for $T$ periods is

$$
L(\beta,\pi,F_{\epsilon};x_{it},w_{it}) = \Pi_{i=1}^N \Pi_{t=1}^T \Pr[d_{it}=1|\Omega{it}]^{d_{it}} \Pr[d_{it}=0|\Omega{it}]^{1-d_{it}}
$$

**Notice** that we are treating this as a single cross-section at this point. There is no dynamics in the model up to now.

# Simulating a data set

We start out with a simulated data set that we will use in estimation.

```{r,echo=FALSE,message=FALSE,warning=FALSE}
# true parameters
p = list()
p$beta1 = 2.5
p$beta2 = 0.5
p$pi = 1.5
p$gamma = 0.6
p$N = 100
p$T = 60  # final year of observation

get.data <- function(p){
  d = data.table(expand.grid(id = 1:p$N,it=1:p$T))
  d[,n := sample(c(0,1,2),size=1), by=id]
  d[,x1 := rnorm(n=nrow(d))]
  d[,x2 := rnorm(n=nrow(d))]
  d[,w := rnorm(nrow(d),mean=3)]
  d[,eps := rnorm(mean=0,sd=1,n=nrow(d))]
  d[,v:= w - p$pi*n - p$beta1*x1- p$beta2*x2]
  d[,work := eps < v]
}
d <- get.data(p)
d
```


## R-golfing

write a one-liner that computes the means of x,w,eps,v and work from this data.table. count how many characters you had to type with `nchar("YOUR_CODE_IN_QUOTES_LIKE_THIS")`. I have 32.

```{r}
nchar("d[,lapply(.SD,mean),.SDcols=4:9]")
d[,lapply(.SD,mean),.SDcols=4:9]
```

1. Why is the mean of w not 2?




### Estimating a static model

What does the likelihood function above remind you of?

Sometimes people use a linear probability model to approximate a probit model. Which parameters can we recover using an LPM here? How well does it predict the average probability of work? Give the quartiles of the probabilities.

```{r}
lmod = lm(work ~w+ x1 + x2 + +n,d)
d[,lpm := predict(lmod,d) ]
summary(lmod)
d[,list(mean(work),mean(lpm))]
d[,quantile(lpm)]
```

What about a probit model? 

```{r}
gmod = glm(work ~ w+ x1 + x2 +n,family =binomial(link="probit"),d)
d[,glm := predict(gmod,d,type="response")]
d[,list(mean(work),mean(lpm),mean(glm))]
summary(gmod)
```

Let's make the wage a bit more realistic. Suppose we knew completed years of education, and we had a variable $z$ capturing *potential experience* (age - educ - 6), which determines the wage, and $\eta$ a random error:

$$
w_{it} = \gamma z_{it} + \eta_{it}
$$

```{r}
get.data <- function(p){
  d = data.table(expand.grid(id = 1:p$N,it=20:p$T))
  d[,n := sample(c(0,1,2),size=1), by=id]
  d[,x1 := rnorm(n=nrow(d))]
  d[,x2 := rnorm(n=nrow(d))]
  d[,educ := sample(10:18,size=1),by=id]
  d[,z := it - educ]
  d[,eta := rnorm(n=nrow(d))]
  d[,w := p$gamma*z + eta]
  d[,eps := rnorm(mean=0,sd=1,n=nrow(d))]
  d[,v:= w - p$pi*n - p$beta1*x1- p$beta2*x2]
  d[,work := eps < v]
}
```

With this modification, our model changes. We have

$$
\begin{align}
v_{it}(x_{it},z_{it},n_{it},\epsilon_{it},\eta_{it}) & = \gamma z_{it} - \pi n_{it} -x_{it}\beta + \eta_{it} - \epsilon_{it} \\
& = \xi_{it}^*(\Omega_{it}) + \xi_{it}
\end{align}
$$
where $\xi = \eta-\epsilon$ is your new composite iid error, and $\xi_{it}^*(\Omega_{it}) = \gamma z_{it} - \pi n_{it} -x_{it}\beta $
Clearly $\Pr[d_{it}=0|\Omega{it}] = 1- \Pr[d_{it}=1|\Omega{it}]$, and so the likelihood for a random sample of $N$ females observed for $T$ periods is

$$
L(\beta,\pi,F_{\epsilon};x_{it},w_{it}) = \Pi_{i=1}^N \Pi_{t=1}^T \Pr[d_{it}=1,w_{it}|\Omega{it}]^{d_{it}} \Pr[d_{it}=0|\Omega{it}]^{1-d_{it}}
$$
where again we have 

$$
\Pr[d_{it}=1,w_{it}|\Omega{it}] = \Pr[d_{it}=1,\eta_{it} = w_{it} - \gamma z_{it}]
$$


<span class="label label-success">Question 7</span> Report the true values, the non biased corrected and the bias corrected. 






